Visual narratives of the COVID-19 pandemic

### 2022-10-27 JDSSV_manuscript_submitted.Rnw
copyediting changes

In other areas, such as visualization of uncertainty, the pandemic presents new challenges for scientific communication. How do we effectively show the uncertainty in the data amid flawed tests, new strains, changing diagnostic criteria, and incredible variability in global public health infrastructure? Moreover, how do we do this without further decreasing public trust in institutions responsible for shepharding us through the pandemic? While most designers chose to side-step these questions in the moment, as we assess the pandemic in graphical form, we should also consider what was omitted from our visual renderings.

@@

\subsection{Data Visualization Design and Testing}
Having briefly discussed the media landscape surrounding the COVID pandemic and goals for communicating about data effectively, we should also briefly discuss the choices that go into creating visual representations of data intended to convey quantitative information.
While the overall goal of a data visualization is to tell the story of the data clearly and effectively, how this story is conveyed involves a series of design choices: which variables are shown, what limits are imposed, what visual forms are used to represent the data? 
How will the chart (and any surrounding story, in journalism) be presented to the viewer? 
As the designer makes these choices, they have in mind an audience, and the audience's assumed characteristics affect the design choices as well: language, education level, likely medium used to access the visualization (computer, physical newspaper, tablet, cell phone) - all of these considerations also impact the design of the chart and what ideas the designer considers during the creative process.

Once the basic form of the chart is determined, there are additional considerations that may be specified by the stylistic conventions of different news organizations, but may also be deliberately chosen for visual effect: color palettes, fonts, spatial layouts, and annotations may be used to highlight information, convey additional points of a narrative, or provide helpful information about how to read an unfamiliar chart.
While each of these design choices could be (and have been) the subject of  additional paper(s), it is important to acknowledge that the graphics we analyze in this paper did not simply pop into being - they are the product of a series of decisions, some consciously made and others imposed by convention or technological capabilities.
We will highlight particular choices which were effective (or not) throughout this paper, but we will also leave some undiscussed for the sake of brevity.

@@

\begin{enumerate}
\item Absolute value unsuitability: As explained in \citep{monmonier2005, slocum2008, speckmann2010} among others, choropleth maps are fundamentally unsuitable for the representation of absolute numbers. Viewers tend to integrate similarly coloured areas of the map unconsciously, and perceive choropleths as representations of density. They also do not help to convey the desired message as the absolute numbers of COVID-19 cases are strongly influenced by the population size of the country as well as by the number of tests performed and the accuracy of the recording and reporting system. 
\item The area-bias: The visual impression is determined more by the colour and the geographical area of the individual countries than by the number of COVID cases. Since the countries of the world differ extremely in area, the visual assessment is distorted, especially in the case of neighbouring countries with similar numbers but different areas.
\item Colour-scheme selection: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997, color2021}. The choice of a continuous scale or a categorical scale, the choice of a scale that promotes the recognition of patterns or a scale that supports the filtering out of specific map details, influences the quality of choropleth maps.
\end{enumerate}

@@

Another modification of the standard choropleth to address the area-bias problem is to relax the spatial borders of a region in order to show magnitude using area (as well as color). Various algorithms can be used to modify geographic boundaries to maintain spatial continuity, or shapes such as hexagons or squares can be used to represent geographic entities (as in \Cref{fig:hex-cartogram}); with either choice, geometry, topology, or both must be sacrificed.  While these methods are more popular in the academic visualisation literature, including analysis of the spread of COVID-19\citep{yalcinMappingGlobalSpatiotemporal2022}, they were not overly common in the media for global representations of COVID information. This may be because cartograms require more visual effort on the part of the viewer; in addition, news organizations may be reluctant to assume their readers have the geographic sophistication necessary to translate between the distorted map of the globe as drawn on the cartogram and the undistorted mental map of the globe used for mental comparison purposes. 
A spatial dot density map might have allowed for analysis of data across geographic boundaries, but this approach was not commonly used to show global case counts, though it was highlighted as a useful option at the national level \citep{fieldMappingCoronavirusResponsibly}, as shown in \Cref{fig:china-dot-map}.

@@

Throughout the pandemic, both in choropleth maps and time-series graphs, creators have had to grapple with scale. When scales automatically adjust, mental comparisons across time can become misleading, because the colour scale from yesterday does not mean the same thing as it did today, as the range has expanded and the colours have stayed the same. 
It is not that often that day-to-day design changes (or lack thereof) cause fundamental problems with the interpretation of data visualisation, but during the COVID 19 pandemic this situation was both extremely common and had real consequences: people would glance at the current COVID risk map and assume that "red" meant the same thing that it did yesterday, despite the fact that the underlying values had changed\citep{abiadColorDiagramWHOLebanon2020,rubelHeyUpshotNYTCurious2020,matthewsLessonFutzingData2020}. 
One example of this issue over two months of NY Times graphics is shown in \Cref{fig:sparklines-heatmap-nyt}; the overall color scheme used changes once, but the variable displayed changes (at least) 4 times over the course of a two month period.

This problem is more noticeable in choropleths and other charts which show case counts based on color scale alone than in time-series charts, where we are more likely to expect and notice a scale change, but it is still a fundamental issue across chart types and of particular concern when there is no natural upper bound on the variable displayed. 
Some designers leaned into this by arranging their graphics to highlight the fact that cases were ``off the charts", as in \Cref{fig:jbm-tweet}, implicitly cuing readers that the situation was exceptional and worthy of their attention.

@@

Early in the pandemic, as case numbers grew exponentially without appropriate public health measures to restrain them, the scale problem for choropleth maps was particularly complex: over time either the colour mapping changes or the map becomes useless due to compression against the upper limits of the scale, as in \citet{panchadsaramIfWeAre2020}. 
The magnitude of the problem also depends on whether the scale uses a set of bins (a discrete mapping to a continuous variable) or uses a continuous color mapping; in some cases, outlets switched between the two options over time or changed the underlying unit entirely. 
Interestingly, media outlets and municipal dashboards were subject to criticism when the color scales changed in response to increasing case counts\citep{boeckDataVizDecisionmakingAction2020} as well as when the color scale did not change\citep{sandalowChiTribGraphicsPleaseUpdate2020}, illustrating the difficulty of designing visualizations amid competing constraints and a constantly evolving pandemic.

@@

The most obvious time-series plot is also the most common: plotting cases against date, often with a 7 or 14-day smooth to handle periodicity in case reporting due to the work week, as shown in \Cref{fig:wapo-COVID-time-series}. This allows for individuals to assess the current situation and easily compare the current number of cases to past times for which the individual has a direct reference for what to expect in impact to daily life. Often, as in \Cref{fig:wapo-COVID-time-series}, these time series show multiple variables: cases, deaths, and hospitalizations (and sometimes vaccinations) to provide a more comprehensive view of the situation; this is particularly useful as there has been some decoupling of case numbers and hospitalizations/deaths with the emergence of effective vaccinations and strains which are more contagious but potentially also less severe. Many outlets also provide additional charts designed to provide some immediate context to readers, as shown in \Cref{fig:wapo-context}.

\begin{figure}
\centering
\includegraphics[width=.5\linewidth]{wapo-COVID-context}
\caption{Some outlets provide contextual information to assist readers with interpreting time-series plots \citep{CoronavirusCasesTracking2022}.}
\label{fig:wapo-context}
\end{figure}

However, there are some limitations with this presentation: if we are interested in showing the change in counts over time, viewers must assess the slope of the line, rather than its position. We know that slope judgments are much less accurate than position \citep{clevelandGraphicalPerceptionVisual1987}, even in relatively simple situations; when we add in our ability to assess exponential growth, our perceptual accuracy is even more suspect. This issue will be discussed further in \Cref{sec:logs} when we discuss log scales, which are one solution to our misperception of exponential growth.

@@

Other time-series representations were more abstract, sacrificing exact data representations for a higher-level summary comparing regions and points in time without the burden of numerical calculations. Several versions from the New York Times are shown in \Cref{fig:sparklines-heatmap-nyt}. The advantage of these displays is that they are very simple and allow for viewers to gain an intuitive understanding of the data, however, they do not present precise numerical information and even the color scale is (perhaps intentionally) opaque -- it would be extremely difficult to translate the visualization into any sort of accurate numerical estimate. In addition, by hiding these details from the user, it is very difficult to identify when the underlying mathematical representation changes over time, as it did several times during the summer of 2020. The use of a similar color scheme as well as similar geometric representations made it extremely difficult for users to identify that the value represented had changed. Clearly, the purpose of these charts is not to provide numerical precision, but rather a comparative assessment of the status of one region relative to others as well as relative to previous time points.

@@

\begin{figure}
\centering
\begin{subfigure}[t]{.25\textwidth}
\includegraphics[width=\textwidth]{nyt-wayback-20200616}
\caption{June 16, 2020}\label{fig:nyt-colors1}
\end{subfigure}\hfill
\begin{subfigure}[t]{.25\textwidth}
\includegraphics[width=\textwidth]{nyt-wayback-20200716}
\caption{July 16, 2020}\label{fig:nyt-colors2}
\end{subfigure}\hfill
\begin{subfigure}[t]{.25\textwidth}
\includegraphics[width=\textwidth]{nyt-wayback-20200718}
\caption{July 18, 2020}\label{fig:nyt-colors3}
\end{subfigure}\hfill
\begin{subfigure}[t]{.25\textwidth}
\includegraphics[width=\textwidth]{nyt-wayback-20200814}
\caption{August 1, 2020}\label{fig:nyt-colors4}
\end{subfigure}\hfill
\caption{A series of screenshots of the NY Times state-by-state display showing evolving color schemes and even metrics; initially, a discrete scale of two-week change in cases is shown; then, a similar color scheme is used with a different metric. Around July 17, the color scheme and variable shown changed to weekly cases per capita (from "fewer" to "more"), and by August 14, the variable shown changed back again ("falling" to "rising") while the color scheme stayed the same.}\label{fig:sparklines-heatmap-nyt}
\end{figure}

One unique time-series plot created a representation of the case counts in the frequency domain: instead of plotting cases as they occurred, the chart instead shows line segments from 0 to N in $y$, where $N$ is a predefined number of cases; then the chart resets to show the next angle. This produces a sense of intensity of the waves of the pandemic over time as shown in \Cref{fig:freq-domain}. Interestingly, this frequency representation can also be easily transferred into an audio domain, providing access for visually impaired users as well as a multimodal representation of the data for sighted users.

\begin{figure}
\centering
\includegraphics[width=.75\linewidth]{frequency-domain}
\caption{Frequency-domain case count graphics. Here, each 1000 deaths are represented by a single line segment whose slope represents the time taken to reach the next 1000 deaths. This provides viewers with a sense of the pace of the epidemic, rather than the raw case counts in more standard time-series representations. This representation can also be shown in the auditory domain, providing access to those who are typically excluded from visualizations due to vision loss \citep{vuillemotSonificationCOVID19Data2020}.}
\label{fig:freq-domain}
\end{figure}

There are many different ways that designers can use time-series data to provide additional contextual information and facilitate comparisons. \Cref{fig:ft-waves} shows COVID cases, positivity rate, and hospital admissions from South Africa's Gauteng province, which was one of the first areas to experience a large wave of the Omicron variant. This chart provides context for the Omicron wave, showing that while infections are occurring at a much faster rate than in previous waves, hospital admissions seem to be approximately following previous trends. In addition, it is clear that the chart is highlighting the Omicron variant from the colour selection: the red used to show the Omicron numbers is a sharp contrast from the green and blue shades used to show the first three waves of the pandemic. Notably, even though red and green in combination are usually to be avoided for the sake of colourblind viewers, the shades of blue/green variants which are chosen are light enough that the red stands out even to someone with colour vision deficiency.

@@

While traditionally, time-series information has been limited to charts with time on a linear x-axis, many charts manipulated time to highlight different aspects of the pandemic. In some cases, these modifications provided large amounts of utility for a relatively small increase in complexity: for example, John Burn-Murdoch's decision to index Financial Times COVID graphs to the days since the $N$th case or death. This decision is a small modification of the basic time-series plot, but has important consequences in the chart's readability as well as the visual emphasis for users: the graphs in the Financial Times were designed to emphasize "... that there's an inevitability about how coronavirus spreads... even if there are only a few cases in your country today, based on all the data we have, you will end up going along that path, the same path that the likes of Italy and Spain have been on so tragically."\citep{hannenCoronavirusTrajectoryTracker2020a} By placing all countries on an even footing divorced from the temporal randomness of when the first case appeared, the viewer can effectively compare the effect of government policies and other interventions on the trajectory over time\citep{burn-murdochAllCountriesDeath2020,burn-murdochLatestCaseTrajectories2020}.

@@

Other designers used the capabilities of web graphics to show the time component in the data through animation. In some cases, this took the form of animated bubble plots XXX add XXX, while others dispensed with the traditional form of a line graph altogether, showing rising case counts using a "bar chart race": an animated series of bar charts over time that show how case counts in each country are increasing. A typical example can be found on YouTube\citep{clockCoronavirusCasesBar2022}; a screenshot is shown in \Cref{fig:barchart-race}. This type of chart is effective in showing instantaneous changes in counts over days for countries with an extremely high number of cases, but may make it difficult for the viewer to process how cases are changing in countries that do not dominate case counts. By focusing viewer attention on changes in relative totals of case counts, the graphic tends to hide steady, proportional increases in cases across the globe, making it somewhat difficult to perceive large global trends relative to small changes in comparative case counts.

@@

As people grappled with the scale of the pandemic and the lives lost as a result, a different set of data displays attempted to provide context to this loss over time. In the New York times, this took the form of a story showing the accumulation of the first 100,000 deaths due to COVID in the US, with occasional short quotes from individuals' obituaries to humanize the names\citep{barryRemembering1000002020}. While this is not a typical time-series chart, as the user scrolled through the page the pace of the names increases, showing the scale of the pandemic in a visceral way. A less sentimental dot-density timeline by the New York Times about 9 months later chronicled the nation reaching 450,000 deaths; the difference between the two visualizations brings to mind the quote oft attributed to Stalin: "One death is a tragedy; a million deaths is a statistic" \citep{investigatorSingleDeathTragedy2016}.

A slightly different approach by the BBC's Visual and Data Journalism team, shown in \Cref{fig:bbc-flower}, displayed the COVID cases and deaths as a growing flower, with the stem proportional to the number of cases and the flower petals showing the deaths over time\citep{thebbcvisualanddatajournalismteamCoronavirusHowCan2020}. The visualization comes with sound as well, an attempt to make the pandemic an auditory experience as well as a visual one. The form, animation, and auditory experience are all indicative of an emotional appeal \citep{kostelnick_re-emergence_2016,dignazio_data_2020}; with the flower representation as an explicitly recognized symbol of grief; it is clearly not about showing specific case counts and deaths numerically.

@@

The plague is always the others: This is the short formula for dealing with infectious diseases from a historical perspective \citep{thiessen2021}. Closing the borders and restricting access to the country became a popular means of controlling the spread of the disease at several different points during the pandemic. Unsurprisingly, the prevailing visual narrative focused on comparisons often fueled by political rivalry, historical dependencies, recent withdrawal from supranational institutions or regional competitions. An ongoing debate about the true extent of the dangers of COVID-19 and how best to combat it, combined with daily availability and public access to data across administrative levels, fostered ongoing competition and the use of leaderboards to show how bad case counts were somewhere else.

@@

Often, these comparative charts were combined with policy discussions, with individuals challenged to spot certain policy interventions on the case-count graphs of different localities with different policies\citep{weissThese12Graphs2020}. \Cref{fig:policy-comparison} shows a chart from an article which appeared in The Federalist, a conservative outlet in the United States, suggesting that mask mandates are ineffective because it is difficult to spot the impact of the mask mandate on the overall trend of cases. Of course, mask mandates are but one component of a much broader pandemic management strategy, but the goal of the chart is clear: the reader is supposed to conclude that cases and masks are not associated. These charts were so common on social media that media outlets ran stories identifying them as misinformation \citep{reutersstaffFactCheckMask2020}; clearly, comparative charts were effectively deployed to misinform as well as to inform.

@@

One issue with assessing the use of log scales is that their effectiveness changes with the stage of the pandemic and the amount and variety of data shown. Initially, log scales were incredibly useful at showing case counts, because minimal mitigation measures were in place and the growth of case counts (or presumptive positive cases, in absence of available testing) was fairly close to exponential. In addition, the use of log scales allowed for the comparison of total cases across entities with large population differences: in the US, we could compare cases in New York and California with cases in Michigan and Washington, even though the populations (and corresponding case counts) of Michigan and Washington are much lower than the populations (and case counts) of either New York or California.

@@

% Problems with log scales
However, after the first wave of COVID, the issues with log scales became more apparent: it was difficult to detect slight increases in case counts that indicated the beginning of a new wave amid a background level of spread, as demonstrated in  \Cref{fig:log-scale-failures}. Diagonal reference lines from the origin were also less helpful, as the growth of cases or deaths was no longer approximately exponential and varied over time; for these reference lines to be effective there would need to be a clear idea of when the case counts started to increase exponentially, which is difficult to determine whilst in the thick of a potential COVID wave. Occasionally graphs with manually drawn reference lines were made available, but usually only in a retrospective manner, as in \Cref{fig:log-scale-ref-multiple}.

@@

Unfortunately, there are not a lot of good options for supporting forecasting decisions in mathematically unsophisticated viewers. Guide lines, like those used in the Financial Times, were helpful in the initial phases of the pandemic, but no outlets that we are aware of shifted the lines to show the growth rate of the current peak (as opposed to the initial peak), and due to the issues shown in \Cref{fig:log-scale-failures}, these guide lines may not have been all that successful in any case. Those hoping to create successful time-series charts that allow forecasting are thus left with two options: deal with our inability to forecast exponential growth, or deal with most individuals' inability to translate log scales into practical reality.

@@

The visual narratives of the pandemic which have persisted over time are due in part to the rise of data journalism and interactive graphics platforms which ensure that every outlet has the ability to host engaging and visually appealing graphics; however, while these graphics are nice to look at, not all are equally functional or useful for communication purposes. Thus, analyzing the rich collection of visualizations produced as part of the pandemic provides a unique opportunity to examine how these graphics were (and were not) effective, as well as the design decisions made during their creation. As in any situation, good tools can be used and misused freely; we have seen charts and graphics used to perpetuate misinformation and misleading claims, even though it is far more common for charts and graphs to be used for good - to educate and inform the public.

>>>>

In other areas, such as visualisation of uncertainty, the pandemic presents new challenges for scientific communication. How do we effectively show the uncertainty in the data amid flawed tests, new strains, changing diagnostic criteria, and incredible variability in global public health infrastructure? Moreover, how do we do this without further decreasing public trust in institutions responsible for shepharding us through the pandemic? While most designers chose to side-step these questions in the moment, as we assess the pandemic in graphical form, we should also consider what was omitted from our visual renderings.

@@

\subsection{Data visualisation Design and Testing}
Having briefly discussed the media landscape surrounding the COVID pandemic and goals for communicating about data effectively, we should also briefly discuss the choices that go into creating visual representations of data intended to convey quantitative information.
While the overall goal of a data visualisation is to tell the story of the data clearly and effectively, how this story is conveyed involves a series of design choices: which variables are shown, what limits are imposed, what visual forms are used to represent the data? 
How will the chart (and any surrounding story, in journalism) be presented to the viewer? 
As the designer makes these choices, they have in mind an audience, and the audience's assumed characteristics affect the design choices as well: language, education level, likely medium used to access the visualisation (computer, physical newspaper, tablet, cell phone) - all of these considerations also impact the design of the chart and what ideas the designer considers during the creative process.

Once the basic form of the chart is determined, there are additional considerations that may be specified by the stylistic conventions of different news organizations, but may also be deliberately chosen for visual effect: colour palettes, fonts, spatial layouts, and annotations may be used to highlight information, convey additional points of a narrative, or provide helpful information about how to read an unfamiliar chart.
While each of these design choices could be (and have been) the subject of  additional paper(s), it is important to acknowledge that the graphics we analyze in this paper did not simply pop into being - they are the product of a series of decisions, some consciously made and others imposed by convention or technological capabilities.
We will highlight particular choices which were effective (or not) throughout this paper, but we will also leave some undiscussed for the sake of brevity.

@@

\begin{enumerate}
\item Absolute value unsuitability: As explained in \citep{monmonier2005, slocum2008, speckmann2010} among others, choropleth maps are fundamentally unsuitable for the representation of absolute numbers. Viewers tend to integrate similarly coloured areas of the map unconsciously, and perceive choropleths as representations of density. They also do not help to convey the desired message as the absolute numbers of COVID-19 cases are strongly influenced by the population size of the country as well as by the number of tests performed and the accuracy of the recording and reporting system. 
\item The area-bias: The visual impression is determined more by the colour and the geographical area of the individual countries than by the number of COVID cases. Since the countries of the world differ extremely in area, the visual assessment is distorted, especially in the case of neighbouring countries with similar numbers but different areas.
\item colour-scheme selection: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997, color2021}. The choice of a continuous scale or a categorical scale, the choice of a scale that promotes the recognition of patterns or a scale that supports the filtering out of specific map details, influences the quality of choropleth maps.
\end{enumerate}

@@

Another modification of the standard choropleth to address the area-bias problem is to relax the spatial borders of a region in order to show magnitude using area (as well as colour). Various algorithms can be used to modify geographic boundaries to maintain spatial continuity, or shapes such as hexagons or squares can be used to represent geographic entities (as in \Cref{fig:hex-cartogram}); with either choice, geometry, topology, or both must be sacrificed.  While these methods are more popular in the academic visualisation literature, including analysis of the spread of COVID-19\citep{yalcinMappingGlobalSpatiotemporal2022}, they were not overly common in the media for global representations of COVID information. This may be because cartograms require more visual effort on the part of the viewer; in addition, news organizations may be reluctant to assume their readers have the geographic sophistication necessary to translate between the distorted map of the globe as drawn on the cartogram and the undistorted mental map of the globe used for mental comparison purposes. 
A spatial dot density map might have allowed for analysis of data across geographic boundaries, but this approach was not commonly used to show global case counts, though it was highlighted as a useful option at the national level \citep{fieldMappingCoronavirusResponsibly}, as shown in \Cref{fig:china-dot-map}.

@@

Throughout the pandemic, both in choropleth maps and time series graphs, creators have had to grapple with scale. When scales automatically adjust, mental comparisons across time can become misleading, because the colour scale from yesterday does not mean the same thing as it did today, as the range has expanded and the colours have stayed the same. 
It is not that often that day-to-day design changes (or lack thereof) cause fundamental problems with the interpretation of data visualisation, but during the COVID 19 pandemic this situation was both extremely common and had real consequences: people would glance at the current COVID risk map and assume that "red" meant the same thing that it did yesterday, despite the fact that the underlying values had changed\citep{abiadColorDiagramWHOLebanon2020,rubelHeyUpshotNYTCurious2020,matthewsLessonFutzingData2020}. 
One example of this issue over two months of NY Times graphics is shown in \Cref{fig:sparklines-heatmap-nyt}; the overall colour scheme used changes once, but the variable displayed changes (at least) 4 times over the course of a two month period.

This problem is more noticeable in choropleths and other charts which show case counts based on colour scale alone than in time series charts, where we are more likely to expect and notice a scale change, but it is still a fundamental issue across chart types and of particular concern when there is no natural upper bound on the variable displayed. 
Some designers leaned into this by arranging their graphics to highlight the fact that cases were ``off the charts", as in \Cref{fig:jbm-tweet}, implicitly cuing readers that the situation was exceptional and worthy of their attention.

@@

Early in the pandemic, as case numbers grew exponentially without appropriate public health measures to restrain them, the scale problem for choropleth maps was particularly complex: over time either the colour mapping changes or the map becomes useless due to compression against the upper limits of the scale, as in \citet{panchadsaramIfWeAre2020}. 
The magnitude of the problem also depends on whether the scale uses a set of bins (a discrete mapping to a continuous variable) or uses a continuous colour mapping; in some cases, outlets switched between the two options over time or changed the underlying unit entirely. 
Interestingly, media outlets and municipal dashboards were subject to criticism when the colour scales changed in response to increasing case counts\citep{boeckDataVizDecisionmakingAction2020} as well as when the colour scale did not change\citep{sandalowChiTribGraphicsPleaseUpdate2020}, illustrating the difficulty of designing visualisations amid competing constraints and a constantly evolving pandemic.

@@

The most obvious time series plot is also the most common: plotting cases against date, often with a 7 or 14-day smooth to handle periodicity in case reporting due to the work week, as shown in \Cref{fig:wapo-COVID-time-series}. This allows for individuals to assess the current situation and easily compare the current number of cases to past times for which the individual has a direct reference for what to expect in impact to daily life. Often, as in \Cref{fig:wapo-COVID-time-series}, these time series show multiple variables: cases, deaths, and hospitalizations (and sometimes vaccinations) to provide a more comprehensive view of the situation; this is particularly useful as there has been some decoupling of case numbers and hospitalizations/deaths with the emergence of effective vaccinations and strains which are more contagious but potentially also less severe. Many outlets also provide additional charts designed to provide some immediate context to readers, as shown in \Cref{fig:wapo-context}.

\begin{figure}
\centering
\includegraphics[width=.5\linewidth]{wapo-COVID-context}
\caption{Some outlets provide contextual information to assist readers with interpreting time series plots \citep{CoronavirusCasesTracking2022}.}
\label{fig:wapo-context}
\end{figure}

However, there are some limitations with this presentation; if we are interested in showing the change in counts over time, viewers must assess the slope of the line, rather than its position. We know that slope judgments are much less accurate than position \citep{clevelandGraphicalPerceptionVisual1987}, even in relatively simple situations; when we add in our ability to assess exponential growth, our perceptual accuracy is even more suspect. This issue will be discussed further in \Cref{sec:logs} when we discuss log scales, which are one solution to our misperception of exponential growth.

@@

Other time series representations were more abstract, sacrificing exact data representations for a higher-level summary comparing regions and points in time without the burden of numerical calculations. Several versions from the New York Times are shown in \Cref{fig:sparklines-heatmap-nyt}. The advantage of these displays is that they are very simple and allow for viewers to gain an intuitive understanding of the data, however, they do not present precise numerical information and even the colour scale is (perhaps intentionally) opaque -- it would be extremely difficult to translate the visualisation into any sort of accurate numerical estimate. In addition, by hiding these details from the user, it is very difficult to identify when the underlying mathematical representation changes over time, as it did several times during the summer of 2020. The use of a similar colour scheme as well as similar geometric representations made it extremely difficult for users to identify that the value represented had changed. Clearly, the purpose of these charts is not to provide numerical precision, but rather a comparative assessment of the status of one region relative to others as well as relative to previous time points.

@@

\begin{figure}
\centering
\begin{subfigure}[t]{.25\textwidth}
\includegraphics[width=\textwidth]{nyt-wayback-20200616}
\caption{June 16, 2020}\label{fig:nyt-colors1}
\end{subfigure}\hfill
\begin{subfigure}[t]{.25\textwidth}
\includegraphics[width=\textwidth]{nyt-wayback-20200716}
\caption{July 16, 2020}\label{fig:nyt-colors2}
\end{subfigure}\hfill
\begin{subfigure}[t]{.25\textwidth}
\includegraphics[width=\textwidth]{nyt-wayback-20200718}
\caption{July 18, 2020}\label{fig:nyt-colors3}
\end{subfigure}\hfill
\begin{subfigure}[t]{.25\textwidth}
\includegraphics[width=\textwidth]{nyt-wayback-20200814}
\caption{August 1, 2020}\label{fig:nyt-colors4}
\end{subfigure}\hfill
\caption{A series of screenshots of the NY Times state-by-state display showing evolving colour schemes and even metrics; initially, a discrete scale of two-week change in cases is shown; then, a similar colour scheme is used with a different metric. Around July 17, the colour scheme and variable shown changed to weekly cases per capita (from "fewer" to "more"), and by August 14, the variable shown changed back again ("falling" to "rising") while the colour scheme stayed the same.}\label{fig:sparklines-heatmap-nyt}
\end{figure}

One unique time series plot created a representation of the case counts in the frequency domain; instead of plotting cases as they occurred, the chart instead shows line segments from 0 to N in $y$, where $N$ is a predefined number of cases; then the chart resets to show the next angle. This produces a sense of intensity of the waves of the pandemic over time as shown in \Cref{fig:freq-domain}. Interestingly, this frequency representation can also be easily transferred into an audio domain, providing access for visually impaired users as well as a multimodal representation of the data for sighted users.

\begin{figure}
\centering
\includegraphics[width=.75\linewidth]{frequency-domain}
\caption{Frequency-domain case count graphics. Here, each 1000 deaths are represented by a single line segment whose slope represents the time taken to reach the next 1000 deaths. This provides viewers with a sense of the pace of the epidemic, rather than the raw case counts in more standard time series representations. This representation can also be shown in the auditory domain, providing access to those who are typically excluded from visualisations due to vision loss \citep{vuillemotSonificationCOVID19Data2020}.}
\label{fig:freq-domain}
\end{figure}

There are many different ways that designers can use time series data to provide additional contextual information and facilitate comparisons. \Cref{fig:ft-waves} shows COVID cases, positivity rate, and hospital admissions from South Africa's Gauteng province, which was one of the first areas to experience a large wave of the Omicron variant. This chart provides context for the Omicron wave, showing that while infections are occurring at a much faster rate than in previous waves, hospital admissions seem to be approximately following previous trends. In addition, it is clear that the chart is highlighting the Omicron variant from the colour selection: the red used to show the Omicron numbers is a sharp contrast from the green and blue shades used to show the first three waves of the pandemic. Notably, even though red and green in combination are usually to be avoided for the sake of colourblind viewers, the shades of blue/green variants which are chosen are light enough that the red stands out even to someone with colour vision deficiency.

@@

While traditionally, time series information has been limited to charts with time on a linear x-axis, many charts manipulated time to highlight different aspects of the pandemic. In some cases, these modifications provided large amounts of utility for a relatively small increase in complexity; for example, John Burn-Murdoch's decision to index Financial Times COVID graphs to the days since the $N$th case or death. This decision is a small modification of the basic time series plot, but has important consequences in the chart's readability as well as the visual emphasis for users: the graphs in the Financial Times were designed to emphasize "... that there's an inevitability about how coronavirus spreads... even if there are only a few cases in your country today, based on all the data we have, you will end up going along that path, the same path that the likes of Italy and Spain have been on so tragically."\citep{hannenCoronavirusTrajectoryTracker2020a} By placing all countries on an even footing divorced from the temporal randomness of when the first case appeared, the viewer can effectively compare the effect of government policies and other interventions on the trajectory over time\citep{burn-murdochAllCountriesDeath2020,burn-murdochLatestCaseTrajectories2020}.

@@

Other designers used the capabilities of web graphics to show the time component in the data through animation. In some cases, this took the form of animated bubble plots XXX add XXX, while others dispensed with the traditional form of a line graph altogether, showing rising case counts using a "bar chart race"; an animated series of bar charts over time that show how case counts in each country are increasing. A typical example can be found on YouTube\citep{clockCoronavirusCasesBar2022}; a screenshot is shown in \Cref{fig:barchart-race}. This type of chart is effective in showing instantaneous changes in counts over days for countries with an extremely high number of cases, but may make it difficult for the viewer to process how cases are changing in countries that do not dominate case counts. By focusing viewer attention on changes in relative totals of case counts, the graphic tends to hide steady, proportional increases in cases across the globe, making it somewhat difficult to perceive large global trends relative to small changes in comparative case counts.

@@

As people grappled with the scale of the pandemic and the lives lost as a result, a different set of data displays attempted to provide context to this loss over time. In the New York times, this took the form of a story showing the accumulation of the first 100,000 deaths due to COVID in the US, with occasional short quotes from individuals' obituaries to humanize the names\citep{barryRemembering1000002020}. While this is not a typical time series chart, as the user scrolled through the page the pace of the names increases, showing the scale of the pandemic in a visceral way. A less sentimental dot-density timeline by the New York Times about 9 months later chronicled the nation reaching 450,000 deaths; the difference between the two visualisations brings to mind the quote oft attributed to Stalin: "One death is a tragedy; a million deaths is a statistic" \citep{investigatorSingleDeathTragedy2016}.

A slightly different approach by the BBC's Visual and Data Journalism team, shown in \Cref{fig:bbc-flower}, displayed the COVID cases and deaths as a growing flower, with the stem proportional to the number of cases and the flower petals showing the deaths over time\citep{thebbcvisualanddatajournalismteamCoronavirusHowCan2020}. The visualisation comes with sound as well, an attempt to make the pandemic an auditory experience as well as a visual one. The form, animation, and auditory experience are all indicative of an emotional appeal \citep{kostelnick_re-emergence_2016,dignazio_data_2020}; with the flower representation as an explicitly recognized symbol of grief; it is clearly not about showing specific case counts and deaths numerically.

@@

"The plague is always the others." This is the short formula for dealing with infectious diseases from a historical perspective \citep{thiessen2021}. Closing the borders and restricting access to the country became a popular means of controlling the spread of the disease at several different points during the pandemic. Unsurprisingly, the prevailing visual narrative focused on comparisons often fueled by political rivalry, historical dependencies, recent withdrawal from supranational institutions or regional competitions. An ongoing debate about the true extent of the dangers of COVID-19 and how best to combat it, combined with daily availability and public access to data across administrative levels, fostered ongoing competition and the use of leaderboards to show how bad case counts were somewhere else.

@@

Often, these comparative charts were combined with policy discussions, with individuals challenged to spot certain policy interventions on the case-count graphs of different localities with different policies\citep{weissThese12Graphs2020}. \Cref{fig:policy-comparison} shows a chart from an article which appeared in The Federalist, a conservative outlet in the United States, suggesting that mask mandates are ineffective because it is difficult to spot the impact of the mask mandate on the overall trend of cases. Of course, mask mandates are but one component of a much broader pandemic management strategy, but the goal of the chart is clear; the reader is supposed to conclude that cases and masks are not associated. These charts were so common on social media that media outlets ran stories identifying them as misinformation \citep{reutersstaffFactCheckMask2020}; clearly, comparative charts were effectively deployed to misinform as well as to inform.

@@

One issue with assessing the use of log scales is that their effectiveness changes with the stage of the pandemic and the amount and variety of data shown. Initially, log scales were incredibly useful at showing case counts, because minimal mitigation measures were in place and the growth of case counts (or presumptive positive cases, in absence of available testing) was fairly close to exponential. In addition, the use of log scales allowed for the comparison of total cases across entities with large population differences; in the US, we could compare cases in New York and California with cases in Michigan and Washington, even though the populations (and corresponding case counts) of Michigan and Washington are much lower than the populations (and case counts) of either New York or California.

@@

% Problems with log scales
However, after the first wave of COVID, the issues with log scales became more apparent; it was difficult to detect slight increases in case counts that indicated the beginning of a new wave amid a background level of spread, as demonstrated in  \Cref{fig:log-scale-failures}. Diagonal reference lines from the origin were also less helpful, as the growth of cases or deaths was no longer approximately exponential and varied over time; for these reference lines to be effective there would need to be a clear idea of when the case counts started to increase exponentially, which is difficult to determine whilst in the thick of a potential COVID wave. Occasionally graphs with manually drawn reference lines were made available, but usually only in a retrospective manner, as in \Cref{fig:log-scale-ref-multiple}.

@@

Unfortunately, there are not a lot of good options for supporting forecasting decisions in mathematically unsophisticated viewers. Guide lines, like those used in the Financial Times, were helpful in the initial phases of the pandemic, but no outlets that we are aware of shifted the lines to show the growth rate of the current peak (as opposed to the initial peak), and due to the issues shown in \Cref{fig:log-scale-failures}, these guide lines may not have been all that successful in any case. Those hoping to create successful time series charts that allow forecasting are thus left with two options: deal with our inability to forecast exponential growth, or deal with most individuals' inability to translate log scales into practical reality.

@@

The visual narratives of the pandemic which have persisted over time are due in part to the rise of data journalism and interactive graphics platforms which ensure that every outlet has the ability to host engaging and visually appealing graphics; however, while these graphics are nice to look at, not all are equally functional or useful for communication purposes. Thus, analyzing the rich collection of visualisations produced as part of the pandemic provides a unique opportunity to examine how these graphics were (and were not) effective, as well as the design decisions made during their creation. As in any situation, good tools can be used and misused freely; we have seen charts and graphics used to perpetuate misinformation and misleading claims, even though it is far more common for charts and graphs to be used for good - to educate and inform the public.
